---
title: "RussellLiu_Scratch"
author: "Russell Liu"
date: "2022-10-24"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Question 1: What do you imagine is the reason for log-transforming the protein levels in biomarker-raw.csv? 

```{r}
library(tidyverse)


# get names
var_names <- read_csv('data/biomarker-raw.csv', 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers (good idea??)
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean <- read_csv('data/biomarker-raw.csv', 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados), 
                ~ trim(scale(log10(.x))[, 1], .at = 3))) %>%
  # reorder columns
  select(group, ados, everything())
```

Scratch for log data: 
```{r}
ggplot(biomarker_clean) + geom_histogram(aes(x = CHIP),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = CEBPB),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = NSE),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = PIAS4),bins = 60)
```

```{r}
biomarker_clean <- read_csv('data/biomarker-raw.csv', 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group))
```

Scratch for raw data
```{r}
ggplot(biomarker_clean) + geom_histogram(aes(x = CHIP),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = CEBPB),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = NSE),bins = 60)
ggplot(biomarker_clean) + geom_histogram(aes(x = PIAS4),bins = 60)
```

We can compare the processed data to the raw data. 
We can clearly see processed data have the same scale of x-axis and y-axis. 
By processing data in Log, x and  y axis don't have much difference in values so we can better view the  data relationship. 

Question 3 part a and b
```{r}
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
load('data/biomarker-clean.RData')

```
Split into training and testing sets at the beginning:
```{r}
set.seed(101422)
biomarker_split <- biomarker_clean %>%
  initial_split(prop = 0.7)
```

```{r}
train_data<-training(biomarker_split)
```

```{r}
test_fn <- function(.df){
  t_test(.df, 
         formula = level ~ group,
         order = c('ASD', 'TD'),
         alternative = 'two-sided',
         var.equal = F)
}

ttests_out <- train_data %>%
  select(-ados) %>%
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  nest(data = c(level, group)) %>% 
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  arrange(p_value) %>%
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)
```

Choose a larger n=20
```{r}
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 20) %>%
  pull(protein)

predictors <- train_data %>%
  select(-c(group, ados))

response <- train_data %>% pull(group) %>% factor()


set.seed(101422)
rf_out <- randomForest(x = predictors, 
                       y = response, 
                       ntree = 1000, 
                       importance = T)

rf_out$confusion

```
```{r}
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 30) %>%
  pull(protein)

predictors <- train_data %>%
  select(-c(group, ados))

response <- train_data %>% pull(group) %>% factor()


set.seed(101422)
rf_out <- randomForest(x = predictors, 
                       y = response, 
                       ntree = 1000, 
                       importance = T)

rf_out$confusion

```

We can see  that the number n doesn't matter. It won't affect the error. 
```{r}
# compute importance scores
proteins_s2 <- rf_out$importance %>% 
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  slice_max(MeanDecreaseGini, n = 20) %>%
  pull(protein)

rf_out$importance
```


