---
title: "R Notebook"
output: html_notebook
---

Question 1: What do you imagine is the reason for log-transforming the protein levels in biomarker-raw.csv? (Hint: look at the distribution of raw values for a sample of proteins.)
```{r}
library(tidyverse)


# get names
var_names <- read_csv("/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-raw.csv", 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers (good idea??)
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean <- read_csv("/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-raw.csv", 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # reorder columns
  select(group, ados, everything())

ggplot(biomarker_clean) + geom_histogram(aes(x = CHIP))
ggplot(biomarker_clean) + geom_histogram(aes(x = CEBPB))
ggplot(biomarker_clean) + geom_histogram(aes(x = NSE))
ggplot(biomarker_clean) + geom_histogram(aes(x = PIAS4))
```

```{r}
library(tidyverse)


# get names
var_names <- read_csv("/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-raw.csv", 
                     col_names = F, 
                     n_max = 2, 
                     col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, 
         abbreviation = V2) %>%
  na.omit()

# function for trimming outliers (good idea??)
trim <- function(x, .at){
  x[abs(x) > .at] <- sign(x[abs(x) > .at])*.at
  return(x)
}

# read in data
biomarker_clean <- read_csv("/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-raw.csv", 
         skip = 2,
         col_select = -2L,
         col_names = c('group', 
                       'empty',
                       pull(var_names, abbreviation),
                       'ados'),
         na = c('-', '')) %>%
  filter(!is.na(group)) %>%
  # log transform, center and scale, and trim
  mutate(across(.cols = -c(group, ados), 
                ~ trim(scale(log10(.x))[, 1], .at = 3))) %>%
  # reorder columns
  select(group, ados, everything())

ggplot(biomarker_clean) + geom_histogram(aes(x = CHIP))
ggplot(biomarker_clean) + geom_histogram(aes(x = CEBPB))
ggplot(biomarker_clean) + geom_histogram(aes(x = NSE))
ggplot(biomarker_clean) + geom_histogram(aes(x = PIAS4))
```

We can see that the graphs in the second portion (with the log transformation) are more dispersed and normal looking. I believe that the log transformation is normalizing the data. We can also see that the x-axis for each of the transformed graphs has the same scale now. I think that this will make analysis of the data much more consistent and easier.

Q4. 
split data set BEFORE preforming tests. Carry out as normal with the TRAINING set.
Choose a larger n
fuzzy is including proteins in either of the three models, doesnt have to be in all 3

I am doing part c
```{r}
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
load('/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-clean.RData')
## MULTIPLE TESTING
####################

# function to compute tests
test_fn <- function(.df){
  t_test(.df, 
         formula = level ~ group,
         order = c('ASD', 'TD'),
         alternative = 'two-sided',
         var.equal = F)
}

ttests_out <- biomarker_clean %>%
  # drop ADOS score
  select(-ados) %>%
  # arrange in long format
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  # nest by protein
  nest(data = c(level, group)) %>% 
  # compute t tests
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  # sort by p-value
  arrange(p_value) %>%
  # multiple testing correction
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)

# select significant proteins
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 10) %>%
  pull(protein)

## RANDOM FOREST
##################

# store predictors and response separately
predictors <- biomarker_clean %>%
  select(-c(group, ados))

response <- biomarker_clean %>% pull(group) %>% factor()

# fit RF
set.seed(101422)
rf_out <- randomForest(x = predictors, 
                       y = response, 
                       ntree = 1000, 
                       importance = T)

# check errors
rf_out$confusion

# compute importance scores
proteins_s2 <- rf_out$importance %>% 
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  slice_max(MeanDecreaseGini, n = 10) %>%
  pull(protein)

## LOGISTIC REGRESSION
#######################
# select subset of interest
proteins_sstar <- intersect(proteins_s1, proteins_s2)
# This gives us 4 proteins: DERM, RELT, IgD, and FSTL1
proteins_sstar

biomarker_sstar <- biomarker_clean %>%
  select(group, any_of(proteins_sstar)) %>%
  mutate(class = (group == 'ASD')) %>%
  select(-group)

# partition into training and test set
set.seed(101422)
biomarker_split <- biomarker_sstar %>%
  initial_split(prop = 0.8)

# fit logistic regression model to training set
fit <- glm(class ~ ., 
           data = training(biomarker_split), 
           family = 'binomial')

# evaluate errors on test set
class_metrics <- metric_set(sensitivity, 
                            specificity, 
                            accuracy,
                            roc_auc)

testing(biomarker_split) %>%
  add_predictions(fit, type = 'response') %>%
  class_metrics(estimate = factor(pred > 0.5),
              truth = factor(class), pred,
              event_level = 'second')
```


```{r}
library(tidyverse)
library(infer)
library(randomForest)
library(tidymodels)
library(modelr)
library(yardstick)
library(BaseSet)
load('/Users/shannon/Documents/PSTAT197/biomarkers-group-1/data/biomarker-clean.RData')
## MULTIPLE TESTING
####################

# function to compute tests
test_fn <- function(.df){
  t_test(.df, 
         formula = level ~ group,
         order = c('ASD', 'TD'),
         alternative = 'two-sided',
         var.equal = F)
}

ttests_out <- biomarker_clean %>%
  # drop ADOS score
  select(-ados) %>%
  # arrange in long format
  pivot_longer(-group, 
               names_to = 'protein', 
               values_to = 'level') %>%
  # nest by protein
  nest(data = c(level, group)) %>% 
  # compute t tests
  mutate(ttest = map(data, test_fn)) %>%
  unnest(ttest) %>%
  # sort by p-value
  arrange(p_value) %>%
  # multiple testing correction
  mutate(m = n(),
         hm = log(m) + 1/(2*m) - digamma(1),
         rank = row_number(),
         p.adj = m*hm*p_value/rank)

# select significant proteins
proteins_s1 <- ttests_out %>%
  slice_min(p.adj, n = 10) %>%
  pull(protein)

## RANDOM FOREST
##################

# store predictors and response separately
predictors <- biomarker_clean %>%
  select(-c(group, ados))

response <- biomarker_clean %>% pull(group) %>% factor()

# fit RF
set.seed(101422)
rf_out <- randomForest(x = predictors, 
                       y = response, 
                       ntree = 1000, 
                       importance = T)

# check errors
rf_out$confusion

# compute importance scores
proteins_s2 <- rf_out$importance %>% 
  as_tibble() %>%
  mutate(protein = rownames(rf_out$importance)) %>%
  slice_max(MeanDecreaseGini, n = 10) %>%
  pull(protein)

## LOGISTIC REGRESSION
#######################

set.seed(27)
proteins <- data.frame(sets = c("proteins_s1", "proteins_s2"),
                       elements = c(proteins_s1, proteins_s2),
                       fuzzy = runif(2))
fuzzy <- tidySet(proteins)
proteins_sstar_f <- intersection(fuzzy, sets = c("proteins_s1", "proteins_s2"))
# This gives us 4 proteins: DERM, RELT, IgD, and FSTL1

proteins_sstar_f <- data.frame(proteins_sstar_f)
proteins_sstar_f$elements

biomarker_sstar_2 <- biomarker_clean %>%
  select(group, any_of(proteins_sstar_f$elements)) %>%
  mutate(class = (group == 'ASD')) %>%
  select(-group)

# partition into training and test set
set.seed(101422)
biomarker_split_2 <- biomarker_sstar_2 %>%
  initial_split(prop = 0.8)

# fit logistic regression model to training set
fit_2 <- glm(class ~ ., 
           data = training(biomarker_split_2), 
           family = 'binomial')

# evaluate errors on test set
class_metrics <- metric_set(sensitivity, 
                            specificity, 
                            accuracy,
                            roc_auc)

testing(biomarker_split_2) %>%
  add_predictions(fit_2, type = 'response') %>%
  class_metrics(estimate = factor(pred > 0.5),
              truth = factor(class), pred,
              event_level = 'second')


```
Nothing changed when including a fuzzy intersection instead of a regular one.



